Efficient Coalition Formation for Web Services

### 2013-03-14 papers/coalition/coalition.tex
Changes

\subsection{Cooperative Game Concepts}
Cooperative game is a branch of game theory that studies strategies of self-interested agents in a setting where agents can increase their payoff by binding agreements and cooperating in groups. We let $N$ to be a set of players. Any subset $S$ of $N$ can form a group that we call it a $coalition$. A \emph{coalitional game} is a pair $G = (N, v)$ where $v$ is \emph{characteristic function} which given a coalition is a function $v: 2^N \to \mathbb{R}$ which maps the set of agents of the coalition to a real number v(S), the worth of S. This number usually resembles the output or payoff or performance of these agents working together as a coalition.  If a coalition S forms then it can divide its worth, v(S) in any possible among its members. The payoff vector $x \in \mathbb{R}^S$ is the amount of payoff being distributed among agents of coalition S. The payoff vector satisfies two conditions:

@@

The second condition is $feasibility$, according to equation below the payoff for each agent cannot be more than coalitions total gain. A payoff vector is also $efficient$ if all the payoff obtained by a coalition is distributed amongst coalition members: $\sum_{i \in S} x_i = v(S)$. This definition of characteristic function works in \emph{transferable utility} (TU) settings, where utility (payoff) is transferable from one player to other, in other words players have common currency and a unit of income is worth same for all players \cite{myerson1991game}.

@@

{\bf Definition 1 (Shapley value)} Given a cooperative game (N, v), the Shapley value of player i is given by\cite{shapley_value}: 
\begin{equation}\label{eq:shapley}
\sum_{S \subseteq N \backslash \left\{i\right\} } \frac{|S|! (n-|S|-1)!}{n!} (v(S \cup \left\{i\right\}) - v(S))
\end{equation}

Shapley value is a unique and fair solution concept for payoff distribution between agents. It basically rewards agents with amount of marginal contribution they have to the coalition.

@@

{\bf Definition 2 (Core)} A payoff vector x denoted by C(N,v) is in the core of a coalitional game (N, v) if and only if
\begin{equation}\label{eq:core}
\forall S \subseteq N, \sum_{x_i \in S} x_i \geq v(S)
\end{equation}
The core if is not empty is basically a set of payoff vectors (x) where no subset of players $S^\prime$ could gain more by deviating and making their own coalition $\sum_{i \in S^\prime} x_i \geq v(S^\prime)$. The sum of payoffs to the agents in any sub-coalition S is at least as large as the amount that these agents could earn by forming a coalition on their own. In a sense it is analogue to Nash equilibrium, except that it about deviations by groups of agents. The core is the strongest and most popular solution concept in cooperative game theory, however its computationally very heavy and is not computationally tractable as number of players increase. The core of some real-world problem games may be empty, which means having characteristic function $v$ there might be no possible distribution of payoff assuring stability of subgroups.

{\bf Definition 3 (Convex cooperative games)} A game with characteristic function v(S) is convex if:
\begin{equation}\label{eq:convex}
V(S) + V(T) \leq v(S \cup T) + v (S \cap T), \forall S,T \subseteq N.
\end{equation}

According to a classic result by Shapley \cite{S1971cores} convex games always have a non-empty core. We will use a variation of convexity condition in our algorithm to check whether our coalitions are stable.

\subsubsection*{$\epsilon$-core}\label{s:epsilon}
%\emph{$\epsilon$-Core:}
%\\
When core set is empty, it means no coalition of players can gain anything by deviating, it is the strongest stable condition. An outcome would be unstable if a coalition can benefit even by a small amount from deviating. This is a strong requirement. In some situations, deviations can be costly or agents may have some loyalty to their coalitions or even it can be computationally intractable to find those small benefits. It would only make sense for a coalition to deviate if the gain from a deviation exceeded a particular cost for example the costs of deviation. $\epsilon-Core$ relaxes the notion of the core, and only require that no coalition can benefit significantly or within a constant amount($\epsilon$) by deviating.

@@

Coalition structure formation is the problem of finding best partition of agents into teams, focusing more on optimizing the coalition's ``owner'' payoff. In these settings performance of individual agent is less important than the \emph{social welfare} of the system, which is the sum of the values of all teams. Having characteristic function game G, a coalition structure (CS) is \emph{socially optimal} is CS belongs to set $argmax_{CS \in CS_N}v(CS)$ where v(CS) is sum of the total valuation function of all coalitions inside CS $(v(CS) = \sum_{C \in CS_N}v(CS))$. The outcome of a characteristic function game in coalition structure settings, consists of two parts; first a disjoint partition of players (agents) into coalitions, called a \emph{coalition structure} (CS) and second a \emph{payoff vector} as mentioned in cooperative game solution concepts, which distributes the value of each coalition among its members.

@@

In this section, we present the architecture of different web service community models and formulate how both web services and community masters as rational agents would maximise their payoff.

@@

In this scenario, there are is a community controller by its master web service. The community master has some request rate $(R_{master})$ from users. Web services need to join a community to be able to get tasks from a master web service. Each web service comes with different quality of service parameters and a throughput $(T_{ws})$. Throughput is the average rate of tasks a web service can perform. Therefore, the master web service is providing tasks with rate of $R_{ws}$ and web services perform tasks with average output metrics of $QoS_{ws}$ and a rate or $T_{ws}$. 
In this setting, we define the value of the coalition as follows:
\begin{equation*}
output(C) = \sum_{ws}{(T_{ws} \times QoS_{ws})}
\end{equation*}
\setlength{\abovedisplayshortskip}{2pt}
\begin{equation*}
v(C) = \left\{ 
  \begin{array}{l l}
    output(C) & \quad \text{if $\sum_{ws}{T_{ws}} \leq R_{master}$}\\
    output(C) \times \frac{R_{master}}{\sum_{ws}{T_{ws}}} & \quad \text{if $\sum_{ws}{T_{ws}} > R_{master}$}
  \end{array} \right.
\end{equation*}

The output of a coalition of web services is the rate and quality of task they can perform. If the rate is more than master's input task rate, it means the web services inside the community are capable of doing more tasks than exists. Considering this the valuation function will balance the output performance and will be the exact amount and quality of work they can perform within the particular community. %** In the case where the limited tasks are distributed among web services uniformly, the value of coalition would be the proportion of the average QoS times their throughput to rate of available requests. **

@@

The membership decision is made based on throughput and QoS of the requested web service. The goal is to have quality web services in community so the community stays stable and no other web services would have incentives to deviate and join ``other'' coalitions. Therefore, we it check the core membership of the coalition of all current community members and the new web service. Our algorithm uses the Shapley value distribution method as described in equation \ref{eq:shapley} to distribute gain of $v(C)$ among all members and then checks if Shapley value payoff vector for this community having characteristic function $v(C)$ is Core stable or not. In Shapley value payoff vector, the payoff for each web service i, is calculated based on their contribution $v(C \cup {i}) - v(C)$ and then averaging over the possible different permutations in which the coalition can be formed, this is why Shapley value distribution is fair. It is proven in convex games the Shapley value lies in the core \cite{DBLP:conf/ijcai/GrecoMPS11, myerson1991game}. Therefore if the Core is non-empty, the Shapley payoff vector is a member of Core.
% so in our algorithm we check the core membership of this payoff vector.

@@

Our algorithm works as follows. We have an established community with a master and some (slave) web services already residing in community. Now we have a request from a web service to join our community. The ideal solution in our method would be analyzing the core or $\epsilon-core$ stability of group having this new member. However the normal core membership algorithms are intractable. We exploit the equation \ref{eq:convex_snow} to check convexity of our characteristic function game having the new member. In that equation we set T to be our community members (C) before having the new web service. We set ${i}$ the new web service, and then verify the equation for S, setting $ S = T / W $ where $W$ is all 1-member subset of set C. We can relax the equation a bit by adding a constant epsilon to left side of equation. We call this method \emph{Depth-1 Convex-Checker} algorithm. If the equation is true for all $W$ we let the member join our community, since he will contribute good enough to make our new community stable. The run time complexity of this algorithm is O(n) which is huge reduction from $O(2^n)$ checking all possible subsets of N. In our second method, we use the same algorithm however this time we set W to be all possible subsets of size two and one of C. We call this method \emph{Depth-2 Convex-Checker} and its run time complexity is $O(n^2)$. It is possible to develop an anytime algorithm by continuing verifying this condition against all subsets of size 3 to N, and stop whenever we are interrupted.

@@

%$CC^{m}$ is \emph{Collabrative Coefficient} for each QoS metric (non-functional parameter), it determines how web services providing that QoS metric individually will be effected in a coalition environment. If it is more than 1, is means they increase performance by working together and they are \emph{super additive}, otherwise it depicts they would perform better in linear sense if they worked alone. 
$w_m$ is the weight of important for a QoS metric, it gives the community the flexibility to emphasize or neglect a specific QoS parameter. Note that $ \sum_{m}{w_m} = 1 $. In this scenario we have no master web services, we assume they either dynamically vote for one, or collaborate and agree on decisions and distribute task in round robin or fair fashion. Here, we utilized the same merge-and-split algorithm which we applied for the \emph{``Web Services and Many Communities''} scenario, to find stable coalitions of web serivces.

@@

In this section, we discuss the experiments we performed for our scenarios to validate applicability and performance of our proposed methods in realistic environments. We created a pool of web services and populated most of their QoS parameters from a real world web service dataset \cite{DBLP:conf/smc/Al-MasriM09a}. We programmed the simulations using Java and executed the simulations on a Intel Xeon X3450 machine with 4GBs of memory. For other parameters we used normal distribution with parameters that make most sense in real world examples.

@@

Figure \ref{performanceall} depicts the results of simulation of optimal $\epsilon-Core$ with 10\% of total community value deviation allowed (one grand community with many web services), \emph{Depth-1 Convex-Checker}, \emph{Depth-2 Convex-Checker}, 3-Way Satisfactory\cite{DBLP:conf/IEEEscc/LimTMB12} and \emph{2 Player Non-Cooperative} methods. The the \emph{optimal $\epsilon$-core} method we capped the coalition size to 24 web services, since anything more than that would require analysis with time complexity of $O(2^{24-1})$ which would made that impractical to run in our simulations. In other methods there were no cap on size of the community and we had communities of size 60 web services at some points. The results show depth-2 and 1 convex checker methods are performing well compared to other methods and all not far from optimal.

@@

As mentioned in preliminaries section \ref{s:preliminaries}, core assumes no coalition of players can gain anything by deviating which is a fairly strong requirement, that is why the notion of $\epsilon-Core$ was introduced. Least-Core $e(G)$ of a game $G$, is the minimum amount of $\epsilon$ in which the core is not empty. We evaluated the least-core value using our valuation function and our set of web services. We pick random number of web services from our data set and formed around 10,000 random coalitions consisting of 3 to 25 web services. We choose 25 as maximum number of members in our grand coalition since it is computationally very complex for bigger coalitions to compute \emph{$\epsilon$}-core. We set $\epsilon$ to a ratio of total payoff of coalition v(C) between 0 and 1. The results in figure \ref{f_leastcore} illustrates, almost 10\% of our random web service coalitions have non-empty core solution and \emph{$\epsilon$}-core solution is always non-empty when we let agents gain only 30\% (of $v(C)$) more, by deviating.

>>>>

\subsection{Cooperative Game Concepts}
Cooperative game is a branch of game theory that studies strategies of self-interested agents in a setting where agents can increase their payoff by binding agreements and cooperating in groups. We let $N$ to be a set of players. Any subset $S$ of $N$ can form a group that we call it a $coalition$. A \emph{coalitional game} is a pair $G = (N, v)$ where $v$ is called a \emph{characteristic function}, which given a coalition, is a function $v: 2^N \to \mathbb{R}$ mapping the set of players of the coalition to a real number $v(S)$, the worth of $S$. This number usually resembles the output or payoff or performance of these players working together as a coalition.  If a coalition $S$ forms then it can divide its worth, $v(S)$ in any possible way among its members. The payoff vector $x \in \mathbb{R}^S$ is the amount of payoff being distributed among agents of coalition $S$. The payoff vector satisfies two conditions:

@@

The second criteria is the \emph{feasibility} condition, according to the equation, the payoff for each agent cannot be more than coalitions total gain. A payoff vector is also \emph{efficient} if all the payoff obtained by a coalition is distributed amongst coalition members: $\sum_{i \in S} x_i = v(S)$. This definition of characteristic function works in \emph{transferable utility} (TU) settings, where utility (payoff) is transferable from one player to other, in other words players have common currency and a unit of income is worth same for all players \cite{myerson1991game}.

@@

{\bf Definition 1 (Shapley value)} Given a cooperative game (N, v), the \emph{Shapley value} of player $i$ is given by\cite{shapley_value}: 
\begin{equation}\label{eq:shapley}
\phi_i(v) = \sum_{S \subseteq N \backslash \left\{i\right\} } \frac{|S|! (n-|S|-1)!}{n!} (v(S \cup \left\{i\right\}) - v(S))
\end{equation}

\emph{Shapley value} is a unique and fair solution concept for payoff distribution between agents. It basically rewards agents with the amount of marginal contribution they have to the coalition.

@@

{\bf Definition 2 (Core)} A payoff vector $\mathbf{x}$, denoted by $C(N,v)$, is in the $core$ of a coalitional game $(N, v)$ if and only if
\begin{equation}\label{eq:core}
\forall S \subseteq N, \sum_{x_i \in S} x_i \geq v(S)
\end{equation}

The core is basically a set of payoff vectors $(\mathbf{x})$ where no subset of players $S^\prime$ could gain more than their current payoff, by deviating and making their own coalition $\sum_{i \in S^\prime} x_i \geq v(S^\prime)$. The sum of payoffs to the players in any sub-coalition $S$ is at least as large as the amount that these players could earn by forming a coalition on their own. In a sense it is analogue to Nash equilibrium, except that it about deviations by groups of agents. The core is the strongest and most popular solution concept in cooperative game theory, however it is a combinatorial problem and is not computationally tractable as number of players increase. The core of some real-world problem games may be empty, which means having characteristic function game of $(N,v)$, there might be no possible distribution of payoff assuring stability of subgroups.

{\bf Definition 3 (Convex cooperative games)} A game with characteristic function $v(S)$ is convex if:
\begin{equation}\label{eq:convex}
V(S) + V(T) \leq v(S \cup T) + v (S \cap T), \forall S,T \subseteq N.
\end{equation}

According to a classic result by Shapley \cite{S1971cores}, convex games always have a non-empty core. We will use a variation of convexity condition in our algorithm to check whether our coalitions are stable.

\subsubsection*{$\epsilon$-core}\label{s:epsilon}
%\emph{$\epsilon$-Core:}
%\\
When \emph{core} set of a game is empty, it means no coalition of players can gain anything by deviating. An outcome would be unstable if a coalition can benefit even by a small amount from deviating. This is a strong requirement. In some situations, deviations can be costly or players may have loyalty to their coalitions, or even it can be computationally intractable to find those small benefits. It would only make sense for a coalition to deviate if the gain from a deviation exceeded a particular cost for example the cost of deviation. \emph{$\epsilon$-core} relaxes the notion of the core, and only requires that, no coalition would benefit significantly, or within a constant amount($\epsilon$) by deviating.

@@

Coalition structure formation is the problem of finding best partition of agents into teams. In these settings, performance of individual agent is less important than the \emph{social welfare} of the system, which is the sum of the values of all teams. Having characteristic function game $G(N,v)$, a coalition structure $(CS)$ is \emph{socially optimal} if $CS$ belongs to set $argmax_{CS \in CS_N}v(CS)$ where $v(CS)$ is sum of the values of all coalitions inside $CS$. $v(CS) = \sum_{C \in CS_N}v(CS)$. 
%The outcome of a characteristic function game in coalition structure settings, consists of two parts; first a disjoint partition of players (agents) into coalitions, called a \emph{coalition structure} (CS) and second a \emph{payoff vector} as mentioned in cooperative game solution concepts, which distributes the value of each coalition among its members.

@@

In this section, we present the architecture of different web service community models and formulate how both web services and community masters as rational agents, would adopt strategies to maximize their payoff.

@@

In this scenario, there is a community controller by its master web service. The community master has a task request rate $(R_{master})$ from users. Web services need to join a community to be able to get tasks from a master web service. Each web service comes with different quality of service parameters and a throughput $(T_{ws})$. Throughput is the average rate of tasks a web service can perform per time unit. Therefore, the master web service is providing tasks with rate of $R_{ws}$ and web services perform tasks with average output quality of $QoS_{ws}$ and a rate or $T_{ws}$. 
In this setting, we define the value of the coalition of the web services within a community as follows:
\begin{equation*}
output(C) = \sum_{ws}{(T_{ws} \times QoS_{ws})}
\end{equation*}
\setlength{\abovedisplayshortskip}{2pt}
\begin{equation*}
v(C) = \left\{ 
  \begin{array}{l l}
    output(C) & \quad \text{if $\sum_{ws}{T_{ws}} \leq R_{master}$}\\
    output(C) \times \frac{R_{master}}{\sum_{ws}{T_{ws}}} & \quad \text{if $\sum_{ws}{T_{ws}} > R_{master}$}
  \end{array} \right.
\end{equation*}

The output of a coalition of web services is the rate and quality of task they can perform. If the rate is more than master's input task rate, it means the web services inside the community are capable of doing more tasks than exists. Considering this the valuation function will balance the output performance and will be the exact rate and quality of work they can perform within the particular community. %** In the case where the limited tasks are distributed among web services uniformly, the value of coalition would be the proportion of the average QoS times their throughput to rate of available requests. **

@@

The membership decision is made based on throughput and \emph{QoS} parameters of the requested web service. The goal is to have quality web services in community so the community stays stable and no other web services would have incentives to deviate and leave the coalitions. Therefore, a basic method would be to check the core membership of the coalition of all current community members (all web services already residing within the community) and the new web service. This algorithm would use the \emph{Shapley value} distribution method as described in equation \ref{eq:shapley} to distribute gain of $v(C)$ among all members and then checks if \emph{Shapley value} payoff vector for this community having characteristic function $v(C)$ is \emph{core} stable or not. In \emph{Shapley value} payoff vector, the payoff for each web service $ws_i$, is calculated based on their marginal contribution $v(C \cup {i}) - v(C)$ over all the possible different permutations in which the coalition can be formed, this is why \emph{Shapley value} distribution is fair. Because of going through all the possible different permutations of subsets of $N$, the nature of the \emph{Shapley value} is combinatorial. This makes it impractical to use as size of our coalitions grow.

It is proven, in convex games the \emph{Shapley value} lies in the core \cite{DBLP:conf/ijcai/GrecoMPS11, myerson1991game}. Therefore if the \emph{Core} is non-empty, the \emph{Shapley value} payoff vector is a member of \emph{Core}.
% so in our algorithm we check the core membership of this payoff vector.

@@

Our algorithm works as follows. We have an established community with a master and some (slave) web services already residing in community. A web service would send a join request to the community. The ideal solution would be analyzing the \emph{core} or \emph{$\epsilon$-core} stability of group having this new member. However the normal core membership algorithms are computationally intractable. We exploit the equation \ref{eq:convex_snow} to check convexity of our characteristic function game having the new member. In the equation, we set $C$ to be our community members $(C)$ before having the new web service. We assign ${i}$ to the new web service, and then verify the equation for $S$, setting $ S = T / W $ where $W$ is all possible 1-member subsets of set $N$. We can relax the equation a bit by adding a constant epsilon to left side of equation. We call this method \emph{Depth-1 Convex-Checker} algorithm. If the equation is true for all $W$ we let the new web service join our community, since the web service will contribute good enough to make our new community stable. The run time complexity of this algorithm is $O(n)$ which is a great reduction from $O(2^n)$ checking all possible subsets of $N$. In our second method, we use the same algorithm however this time we set $W$ to be all possible subsets of size two and one of the community $C$. We call this method \emph{Depth-2 Convex-Checker} and its run time complexity is $O(n^2)$. It is possible to develop an anytime algorithm by continuing verifying this condition against all subsets of size 3 to $|N|$, and stop whenever we the algorithm is interrupted.

@@

%$CC^{m}$ is \emph{Collabrative Coefficient} for each QoS metric (non-functional parameter), it determines how web services providing that QoS metric individually will be effected in a coalition environment. If it is more than 1, is means they increase performance by working together and they are \emph{super additive}, otherwise it depicts they would perform better in linear sense if they worked alone. 
$w_m$ is the weight of important for a \emph{QoS} metric, it gives the community the flexibility to emphasize or neglect a specific \emph{QoS} parameter. Note that $ \sum_{m}{w_m} = 1 $. In this scenario we have no master web services, we assume they either dynamically vote for one, or collaborate and agree on decisions and distribute task in round robin or fair fashion. Here, we utilized the same merge-and-split algorithm which we applied for the \emph{``Web Services and Many Communities''} scenario, to find stable coalitions of web serivces.

@@

In this section, we discuss the experiments we performed for our scenarios to validate applicability and performance of our proposed methods in realistic environments. We created a pool of web services and populated most of their \emph{QoS} parameters from a real world web service dataset \cite{DBLP:conf/smc/Al-MasriM09a}. We programmed the simulations using Java and executed the simulations on a Intel Xeon X3450 machine with 4GBs of memory. For other parameters missing from the dataset, we used normal random distribution with parameters that make most sense in real world examples.

@@

Figure \ref{performanceall} depicts the results of simulation of optimal \emph{$\epsilon$-core} with 10\% of total community value deviation allowed (one grand community with many web services), \emph{Depth-1 Convex-Checker}, \emph{Depth-2 Convex-Checker}, 3-Way Satisfactory\cite{DBLP:conf/IEEEscc/LimTMB12} and \emph{2 Player Non-Cooperative} methods. The the \emph{optimal $\epsilon$-core} method we capped the coalition size to 24 web services, since anything more than that would require analysis with time complexity of $O(2^{24-1})$, making it impractical to run in our simulations. In other methods there were no cap on size of the community and we had communities of size 60 web services at some points. The results show \depth{depth-2 and 1 convex checker} methods are performing well compared to other methods and all not far from optimal.

@@

As mentioned in section \ref{s:preliminaries}, core assumes no coalition of players can gain anything by deviating which is a fairly strong requirement, that is why the notion of \emph{$\epsilon$-core} was introduced. Least-Core $e(G)$ of a game $G$, is the minimum amount of $\epsilon$ in which the core is not empty. We evaluated the least-core value using our valuation function and our set of web services. We pick random number of web services from our data set and formed around 10,000 random coalitions consisting of 3 to 25 web services. We choose 25 as maximum number of members in our grand coalition since it is computationally very complex for bigger coalitions to compute \emph{$\epsilon$-core}. We set $\epsilon$ to a ratio of total payoff of coalition $v(C)$ between 0 and 1. The results in figure \ref{f_leastcore} illustrates, almost 10\% of our random web service coalitions have non-empty core solution and \emph{$\epsilon$-core} solution is always non-empty when we let agents gain only 30\% (of $v(C)$) more, by deviating.
